# Bayesian hierarchical models {#ch-hierarchical}


Usually, experimental data in cognitive science contain "clusters." These are natural groups that contain observations that are more similar within the clusters than between them.
The most common examples of clusters in experimental designs are subjects and experimental items (e.g., words, pictures, objects that are presented to the subjects). These clusters arise because we have multiple (repeated) observations for each subject, and for each item. Such data are sometimes also called dependent data, because the observations within a cluster are not independent. If we want to incorporate this \index{Grouping structure} grouping structure in our analysis, we generally use a \index{Hierarchical model} hierarchical model [also called \index{Multi-level} multi-level or \index{Mixed model} mixed models, @pinheirobates]. This kind of clustering and hierarchical modeling arises as a consequence of the idea of \index{Exchangeability} *exchangeability*.

## Exchangeability and hierarchical models

Exchangeability is the Bayesian analog of the phrase \index{Independent and identically distributed} "independent and identically distributed" that appears regularly in classical (i.e., frequentist) statistics. Some connections and differences between exchangeability and the frequentist concept of independent and identically distributed (iid) are detailed in online section \@ref(app-exch).

Informally, the idea of exchangeability is as follows. Suppose we assign a numerical index to each of the levels of a group (e.g., to each subject). When the levels are exchangeable, we can reassign the indices arbitrarily and lose no information; that is, the joint distribution will remain the same, because we don't have any different prior information for each cluster (here, for each subject). In hierarchical models, we treat the levels of the \index{Group} group as exchangeable, and observations within each level in the group as also exchangeable. We generally include predictors at the level of the observations, those are the predictors that correspond to the experimental manipulations (e.g.,  attentional load, trial number, cloze probability, etc.); and maybe also at the group level, these are predictors that indicate characteristics of the levels in the group (e.g., the working memory capacity score of each subject). Then the conditional distributions given these explanatory variables would be exchangeable; that is, our predictors incorporate all the information that is not exchangeable, and when we factor the predictors out, the observations or units in the group are exchangeable. This is the reason why the item number is an appropriate cluster, but trial number is not: In the first case, if we permute the numbering of the items there is no loss of information because the indexes are exchangeable, all the information about the items is incorporated as predictors in the model. In the second case, the numbering of the trials include information that will be lost if we treat them as exchangeable. For example, consider the case where, as trial numbers increase, subjects get more experienced or fatigued. Even if we are not interested in the specific cluster-level estimates, hierarchical models allow us to generalize to the underlying population (subjects, items) from which the clusters in the sample were drawn. Of course, this generalization is only possible if the samples are representative (e.g., if the subjects and items are randomly drawn); strictly speaking, this condition often does not hold in experimental fields like psychology and linguistics. For more on exchangeability, consult the further reading at the end of the chapter.

Exchangeability is important in Bayesian statistics because of a theorem called the \index{Representation Theorem} Representation Theorem [@deFinetti]. This theorem states that if a sequence of random variables is exchangeable, then the prior distributions on the parameters in a model are a necessary consequence; priors are not merely an arbitrary addition to the frequentist modeling approach that we are familiar with. For a detailed discussion on the Representation Theorem, see chapter 4 of @kendall2004.

Furthermore, exchangeability has been shown [@bernardosmith] to be mathematically equivalent to assuming a hierarchical structure in the model. The argument goes as follows. Suppose that the parameters for each level in a group are $\mu_i$, where the levels are labeled $i=1,\dots,I$. An example of groups is subjects. Suppose also that the data $y_n$, where $n=1,\dots,N$ are observations from these subjects (e.g., pupil size measurements). The data are assumed to be generated as

\begin{equation}
y_n \sim \mathit{Normal}(\mu_{subj[n]},\sigma)
\end{equation}

The notation  $subj[n]$, which roughly follows @GelmanHill2007, identifies the subject index. Suppose that each of 20 subjects provide 50 observations. If the data are ordered by subject id, then $subj[1]$ to $subj[50]$ corresponds to a subject with id $i=1$, $subj[51]$ to $subj[100]$ corresponds to a subject with id $i=2$, and so on.

We can code this representation in a straightforward way in R:

```{r}
N_subj <- 20
N_obs_subj <- 50
N <- N_subj * N_obs_subj
df <- tibble(row = 1:N,
             subj = rep(1:N_subj, each = N_obs_subj))
df
# Example:
c(df$subj[1], df$subj[2], df$subj[51])
```


If the data $y_n$ are exchangeable, the parameters $\mu_i$ are also exchangeable. The fact that the $\mu_i$ are exchangeable can be shown [@bernardosmith]  to be mathematically equivalent to assuming that they come from a common distribution, for example:

\begin{equation}
\mu_i \sim \mathit{Normal}(\mu,\tau)
\end{equation}


To make this more concrete, assume some completely arbitrary true values for the parameters, and generate  observations based on a hierarchical process in R.

```{r, echo = FALSE}
set.seed(123)
```
```{r}
mu <- 100
tau <- 15
sigma <- 4
mu_i <- rnorm(N_subj, mu, tau)
df_h <- mutate(df, y = rnorm(N, mu_i[subj], sigma))
df_h
```


 The parameters $\mu$ and $\tau$, called \index{Hyperparameter} hyperparameters, are unknown and have prior distributions \index{Hyperprior} (hyperpriors) defined for them. This fact leads to a hierarchical relationship between the parameters: there is a common parameter $\mu$ for each of the levels of a group (each of the  subjects), and the parameters $\mu_i$ are assumed to be generated as a function of this common parameter $\mu$. Here, $\tau$ represents \index{Between-group variability} between-group (here, between-subject) variability, and $\sigma$ represents \index{Within-group variability} within-group (within-subject) variability. The three parameters have priors defined for them. The first two priors below are called hyperpriors.

- $p(\mu)$
- $p(\tau)$
- $p(\sigma)$

In such a model, information about $\mu_i$ comes from two sources:

(a) from each of the observed $y_n$ corresponding to the respective $\mu_{subj[n]}$ parameter, and

(b) from the parameters $\mu$ and $\tau$ that led to all the other $y_k$ (where $k\neq n$) being generated.


This is illustrated in Figure \@ref(fig:dags1).

Fit this model in `brms` in the following way. For now, the prior distributions are arbitrary.

```{r, message = FALSE, results = "hide"}
fit_h <- brm(y ~ 1 + (1 | subj), df_h,
             prior =
               c(prior(normal(50, 200), class = Intercept),
                 prior(normal(2, 5), class = sigma),
                 prior(normal(10, 20), class = sd)),
             # increase iterations to avoid convergence issues
             iter = 5000)
```

```{r, eval = FALSE}
fit_h
```

```{r, echo = FALSE}
short_summary(fit_h)
```

In this output, `Intercept` corresponds to the posterior of $\mu$, `sigma` to $\sigma$, and `sd(Intercept)` to $\tau$. There is more information in the `brms` object: we can also get the posteriors for each level of our group (i.e., for each subject). However, rather than estimating $\mu_i$, `brms` estimates the adjustments to $\mu$, $u_i$, named `r_subj[i,Intercept]`, so that $\mu_i = \mu + u_i$. See the code below.

```{r, warning = FALSE}
# Extract the posterior estimates of u_i
u_i_post <- as_draws_df(fit_h) %>%
  select(starts_with("r_subj"))
# Extract the posterior estimate of mu
mu_post <- as_draws_df(fit_h)$b_Intercept
# Build the posterior estimate of mu_i
mu_i_post <- mu_post + u_i_post
colMeans(mu_i_post) %>% unname()
# Compare with true values
mu_i
```

(ref:dags1) A \index{Directed acyclic graph} directed acyclic graph illustrating a hierarchical model \index{Partial pooling} (partial pooling).


```{tikz dags1, fig.cap = "(ref:dags1)", out.width = "100%", echo = FALSE, fig.ext = 'png', cache = FALSE, echo = FALSE, dpi = 1000}
%% hierarchical model:
\begin{tikzpicture}
\matrix[matrix of math nodes, column sep=20pt, row sep=20pt] (mat)
{
          &              &           && \mu, \tau          &       &        &          &        & \\
          & \mu_1        &           &           & \mu_{2} &       & \ldots & \mu_{20} &        &\\
    y_{1} & y_2      & \ldots    & y_{50}    & y_{51}  & \ldots&  \ldots & y_{951} & \ldots & y_{1000}\\
          &              &           & &\sigma              &         &       &         &         &        &\\
};

    % top arrows:
    \draw[->,>=latex] (mat-1-5) -- (mat-2-2);
    \draw[->,>=latex] (mat-1-5) -- (mat-2-5);
    \draw[->,>=latex] (mat-1-5) -- (mat-2-7);
    \draw[->,>=latex] (mat-1-5) -- (mat-2-8);

    % from mu1
    \draw[->,>=latex] (mat-2-2) -- (mat-3-1);
    \draw[->,>=latex] (mat-2-2) -- (mat-3-2);
    \draw[->,>=latex] (mat-2-2) -- (mat-3-3);
    \draw[->,>=latex] (mat-2-2) -- (mat-3-4);

    % from mu2
    \draw[->,>=latex] (mat-2-5) -- (mat-3-5);
    \draw[->,>=latex] (mat-2-5) -- (mat-3-6);
    % from ...
    \draw[->,>=latex] (mat-2-7) -- (mat-3-7);

    \draw[->,>=latex] (mat-2-8) -- (mat-3-8);
    \draw[->,>=latex] (mat-2-8) -- (mat-3-9);
    \draw[->,>=latex] (mat-2-8) -- (mat-3-10);


\foreach \column in {1, 2, 3,4,5,6, 7,8,9,10}
{
    \draw[->,>=latex]  (mat-4-5)-- (mat-3-\column);
}
\end{tikzpicture}
```




There are two other configurations possible that do not involve this hierarchical structure and which represent two alternative, extreme scenarios.

One of these two configurations is called the \index{Complete pooling model} *complete pooling model*, Here, the data $y_n$ are assumed to be generated from a single distribution:

\begin{equation}
y_n \sim \mathit{Normal}(\mu,\sigma).
\end{equation}

This model is an intercept only regression similar to what we saw in chapter \@ref(ch-compbda).

Generate simulated observations in a vector `y` based on arbitrary true values in R in the following way.

```{r}
sigma <- 4
mu <- 100
df_cp <- mutate(df, y = rnorm(N, mu, sigma))
df_cp
```

Fit it in `brms`.

```{r, message = FALSE, results = "hide"}
fit_cp <- brm(y ~ 1, df_cp,
             prior =
               c(prior(normal(50, 200), class = Intercept),
                 prior(normal(2, 5), class = sigma)))
```

```{r, eval = FALSE}
fit_cp
```

```{r, echo = FALSE}
short_summary(fit_cp)
```

The configuration of the complete pooling model is illustrated in Figure \@ref(fig:dags3).

(ref:dags3) A directed acyclic graph illustrating  a complete pooling model.

```{tikz dags3, fig.cap = "(ref:dags3)", out.width = "100%", echo = FALSE, fig.ext = 'png', cache = FALSE, echo = FALSE, dpi = 1000,   out.extra = ''}
%% complete pooling model:
\begin{tikzpicture}
\matrix[matrix of math nodes, column sep=20pt, row sep=20pt] (mat)
{
          &       & \mu   &         \\
    y_{1} & y_{2} & \ldots & y_{1000} \\
           &      & \sigma &        \\
};
\foreach \column in {1, 2,3,4}
{
    \draw[->,>=latex] (mat-1-3) -- (mat-2-\column);
    \draw[->,>=latex] (mat-3-3) -- (mat-2-\column);
}
\end{tikzpicture}
```


The other configuration is called the \index{No pooling model} *no pooling model*; here, each $y_n$ is assumed to be generated from an independent distribution:

\begin{equation}
y_n \sim \mathit{Normal}(\mu_{subj[n]},\sigma)
\end{equation}

```{r, echo = FALSE, eval = FALSE}
dput(round(runif(20, 50, 200)))

```
Generate simulated observations from the no pooling model in R with arbitrary true values.

```{r}
sigma <- 4
mu_i <- c(156, 178, 95, 183, 147, 191, 67, 153, 129, 119, 195,
          150, 172, 97, 110, 115, 78, 126, 175, 80)
df_np <- mutate(df, y = rnorm(N, mu_i[subj], sigma))
df_np
```

Fit the no pooling model in `brms`. By using the formula `0 + factor(subj)`, the common intercept can be removed and the model is forced to estimate one intercept for each `subj` separately. The column `subj` is converted to a factor so that `brms` does not interpret it as a number.

```{r, message = FALSE, results = "hide"}
fit_np <- brm(y ~ 0 + factor(subj), df_np,
             prior =
               c(prior(normal(0, 200), class = b),
                 prior(normal(2, 5), class = sigma)))
```

The summary shows now the 20 estimates of $\mu_i$ as `b_factorsubj` and $\sigma$. (We ignore  `lp__` and `lprior`.)

```{r}
fit_np %>% posterior_summary()
```



Unlike the hierarchical model, now there is no common distribution that generates the $\mu_i$ parameters. This is illustrated in Figure  \@ref(fig:dags2).

(ref:dags2) A directed acyclic graph illustrating a no pooling model.


```{tikz dags2, fig.cap = "(ref:dags2)", out.width = "100%", echo = FALSE, fig.ext = 'png', cache = FALSE, echo = FALSE, dpi = 1000}
%% independent (no pooling) model:
\begin{tikzpicture}
\matrix[matrix of math nodes, column sep=20pt, row sep=20pt] (mat)
{
          &              &           &&           &       &        &          &        & \\
          & \mu_1        &           &           & \mu_{2} &       & \ldots & \mu_{20} &        &\\
    y_{1} & y_2      & \ldots    & y_{50}    & y_{51}  & \ldots&  \ldots & y_{951} & \ldots & y_{1000}\\
          &              &           & &\sigma              &         &       &         &         &        &\\
};


    % from mu1
    \draw[->,>=latex] (mat-2-2) -- (mat-3-1);
    \draw[->,>=latex] (mat-2-2) -- (mat-3-2);
    \draw[->,>=latex] (mat-2-2) -- (mat-3-3);
    \draw[->,>=latex] (mat-2-2) -- (mat-3-4);

    % from mu2
    \draw[->,>=latex] (mat-2-5) -- (mat-3-5);
    \draw[->,>=latex] (mat-2-5) -- (mat-3-6);
    % from ...
    \draw[->,>=latex] (mat-2-7) -- (mat-3-7);

    \draw[->,>=latex] (mat-2-8) -- (mat-3-8);
    \draw[->,>=latex] (mat-2-8) -- (mat-3-9);
    \draw[->,>=latex] (mat-2-8) -- (mat-3-10);


\foreach \column in {1, 2, 3,4,5,6, 7,8,9,10}
{
    \draw[->,>=latex]  (mat-4-5)-- (mat-3-\column);
}
\end{tikzpicture}
```


The hierarchical model lies between these two extremes and for this reason is sometimes called a \index{Partial pooling} *partial pooling model*. One way that the hierarchical model is often described is that the estimates $\mu_i$ "borrow strength" from the parameter $\mu$ (which represents the grand mean in the above example).

An important practical consequence of partial pooling is the idea of "borrowing strength from the mean": if we have very sparse data from a particular member of a group (e.g., missing data from a particular subject), the estimate $\mu_i$ of that particular group member $n$ is determined by the parameter $\mu$. In other words, when the data are sparse for group member $n$, the posterior estimate $\mu_i$ is determined largely by the posterior on $\mu$. In this sense, even the frequentist hierarchical modeling software in R, `lmer()` from the package `lme4`, is essentially Bayesian in formulation (except of course that there is no prior on $\mu$).

So far we have focused on the structure of $\mu$, the location parameter of the likelihood. We could even have partial pooling, complete pooling or no pooling with respect to $\sigma$, the scale parameter of the likelihood. More generally, any parameter of a likelihood can have any of these kinds of pooling.

In the coming sections, we will be looking at each of these models with more detail and using realistic examples.

## A hierarchical model with a normal likelihood: The N400 effect {#sec-N400hierarchical}

\index{Event-related potentials} Event-related potentials \index{ERPs} (ERPs) allow scientists to observe electrophysiological responses in the  brain measured by means of \index{Electroencephalography} electroencephalography \index{EEG} (EEG) that are time-locked to a specific event (i.e., the presentation of the stimuli). A very robust ERP effect in the study of language is the \index{N400} N400. Words with low predictability are accompanied by an *N400 effect* in comparison with high-predictable words, this is a relative negativity that peaks around 300-500 ms after word onset over central parietal scalp sites [first reported in  @kutasReadingSenselessSentences1980;  for semantic anomalies, and in @kutasBrainPotentialsReading1984 for low predictable word; for a review, see @kutasThirtyYearsCounting2011]. The N400 is illustrated in Figure \@ref(fig:N400noun).


```{r, echo = FALSE}
N400_c <- c("Cz", "CP1", "CP2", "P3", "Pz", "P4", "POz")
```

(ref:N400noun) Typical ERPs for the grand average across the N400 spatial window  (central parietal electrodes: `r paste0(N400_c, collapse =", ")`) for high and low \index{Predictability} predictability nouns [specifically from the constraining context of the experiment reported in @NicenboimPreactivation2019]. The x-axis indicates time in seconds and the y-axis indicates voltage in microvolts (unlike many EEG/ERP plots, negative polarity is plotted downwards).

```{r N400noun, results = "hold",fig.height= 3.5, fig.cap =  "(ref:N400noun)", echo = FALSE, warning = FALSE}
noun_s <- readRDS("dataR/external/noun_s.RDS") %>%
  mutate(
    constraint = ifelse(cond %in% c(0, 1),
      "Constraining", "Non-constraining"
    ),
    completion = ifelse(cond %in% c(0, 2), "a", "b"),
    predictability = case_when(
      cond == 0 ~ "high",
      cond == 1 ~ "low",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(
    channel == "neg", .time >= -1.6, .time <= 2.2,
    constraint == "Constraining"
  )

noun_s %>%
  ggplot(aes(x = .time, y = mean_s, linetype = predictability)) +
  labs(linetype = "Predictability", fill = "", color = "", x = "Time (s)", y = "Amplitude (\u03BCV)") +
  scale_x_continuous(limits = c(-.1, .8)) +
  coord_cartesian(xlim = c(-.1, .8), clip = "off") +
  geom_line(size = .5, na.rm = TRUE) +
  geom_vline(xintercept = .3, linetype = "dashed", color = "gray") +
  geom_vline(xintercept = .5, linetype = "dashed", color = "gray")
```


For example, in (1) below, the continuation '*dog*'  has lower predictability than the continuation '*paint*,' and thus we would expect a more negative signal, that is, an N400 effect, in  '*dog*' in (b) in comparison with  '*paint*' in  (a). Predictability is often measured with a cloze task (see section \@ref(sec-binomialcloze)).

1. Example from @kutasBrainPotentialsReading1984
    a. Don't touch the wet paint.
    b. Don't touch the wet dog.


The EEG data are typically recorded in tens of electrodes every couple of milliseconds, but for our purposes (i.e., for  learning about Bayesian hierarchical models), we can safely ignore the complexity of the data. A common way to simplify the high-dimensional EEG data when we are dealing with the N400 is to focus on the average amplitude of the EEG signal at the typical spatio-temporal window of the N400 [for example, see  @frankERPResponseAmount2015].


For this example, we are going to focus on the N400 effect for critical nouns from a subset of the data of @nieuwlandLargescaleReplicationStudy2018. @nieuwlandLargescaleReplicationStudy2018 presented a replication attempt of an original experiment of @delongProbabilisticWordPreactivation2005 with sentences like (2).



2. Example from @delongProbabilisticWordPreactivation2005
    a. The day was breezy so the boy went outside to fly a kite.
    b. The day was breezy so the boy went outside to fly an airplane.


We'll ignore the goal of original experiment [@delongProbabilisticWordPreactivation2005], and its replication [@nieuwlandLargescaleReplicationStudy2018]. We are going to focus on the N400 at the final nouns in the experimental stimuli. In example (2), the final noun '*kite*'  has higher predictability than '*airplane*,' and thus we would expect a more negative signal in  '*airplane*' in (b) in comparison with  '*kite*' in  (a).


To speed up computation,  we restrict the data set of @nieuwlandLargescaleReplicationStudy2018 to the subjects from the Edinburgh lab. This subset of the data can be found in `df_eeg` in the `bcogsci` package.  Center the cloze probability before using it as a predictor.

```{r, echo = FALSE}
# hack:
select <- dplyr::select
```

```{r, message = FALSE}
data("df_eeg")
(df_eeg <- df_eeg %>%
  mutate(c_cloze = cloze - mean(cloze)))
# Number of subjects
df_eeg %>%
  distinct(subj) %>%
  count()
```

One convenient aspect of using averages of EEG data is that they are roughly normally distributed. This allows us to use the normal likelihood. Figure \@ref(fig:histn400) shows the distribution of the data.

```{r histn400, fig.cap="Histogram of the N400 averages for every trial, overlaid is a density plot of a normal distribution.", message=FALSE, fold = TRUE,fig.height = 2.2 }
df_eeg %>% ggplot(aes(n400)) +
  geom_histogram(binwidth = 4,
                 colour = "gray",
                 alpha = .5,
                 aes(y = after_stat(density))) +
  stat_function(fun = dnorm,
                args = list(mean = mean(df_eeg$n400),
                            sd = sd(df_eeg$n400))) +
  xlab("Average voltage in microvolts for
       the N400 spatiotemporal window")
```


### \index{Complete pooling model} Complete pooling model ($M_{cp}$) {#sec-Mcp}

We'll start from the simplest model which is basically the linear regression we encountered in the preceding chapter.

#### Model assumptions

This model, call it $M_{cp}$, makes the following assumptions.

1. The EEG averages for the N400 spatiotemporal window are normally distributed.
2. Observations are independent.
3. There is a linear relationship between cloze and the EEG signal for the trial.

**This model is incorrect for these data due to assumption (2) being violated.**

With the last assumption, we are saying that the difference in the average signal when we compare nouns with cloze probability of 0 and 0.1 is the same as the difference in the signal when we compare nouns with cloze values of 0.1 and 0.2 (or 0.9 and 1). This is just an assumption, and it may not necessarily be the case in the actual data. This means that we are going to get a posterior for $\beta$ conditional on the assumption that the linear relationship holds. Even if the assumption approximately holds, we still don't know how much we deviate from this assumption.

We can now decide on a likelihood and priors.

#### Likelihood and priors

A normal likelihood seems reasonable for these data:

 \begin{equation}
   signal_n \sim \mathit{Normal}( \alpha + c\_cloze_n \cdot \beta,\sigma)
  (\#eq:Mcp)
 \end{equation}

where $n =1, \ldots, N$, and $signal$ is the dependent variable (average signal in the N400 spatiotemporal window in microvolts). The variable $N$ represents the total number of data points.

As always, we need to rely on our previous knowledge and domain expertise to decide on priors. We know that ERPs (signals time-locked to a stimulus) have mean amplitudes of a couple of microvolts: This is easy to see in any plot of the EEG literature. This means that we don't expect the effect of our manipulation to exceed, say, $10 \mu V$. As before, a priori we'll assume that effects can be negative or positive. We can quantify our prior knowledge regarding plausible values of $\beta$ as normally distributed, centered at zero with a standard deviation of $10 \mu V$. (Other values such as  $5 \mu V$ would have been equally reasonable, since this smaller standard deviation would entail that 95% of the prior mass probability is between $-10$ and $10 \mu V$.)

If the signal for each ERP is *baselined*, that is, the mean signal of a time window before the time window of interest is subtracted from the time window of interest, then the mean signal would be relatively close to $0$. Since we know that the ERPs were baselined in this study, we expect that the grand mean of our signal should be relatively close to zero. Our prior for $\alpha$ is then also normally distributed,  and centered at zero with a standard deviation of $10 \mu V$.

The prior for the standard deviation $\sigma$ of our signal distribution requires some thought. We know that EEG signals are quite noisy, and that the standard deviation must be higher than zero. Our prior for  $\sigma$ is a  truncated normal distribution with  location  zero and scale 50. Recall that since we truncate the distribution, the parameters location and scale do not correspond to the mean and standard deviation of the new distribution; see online section \@ref(app-truncation).

We can draw random samples from this truncated distribution and calculate their mean and standard deviation:

```{r}
samples <- rtnorm(20000, mean = 0, sd = 50, a = 0)
c(mean = mean(samples), sd = sd(samples))

```
So we are essentially saying that we assume a priori that we will find the true standard deviation of the signal in the following interval with 95% probability:

```{r}
quantile(samples, probs = c(0.025, 0.975))
# Analytically:
# c(qtnorm(0.025, 0, 50, a = 0), qtnorm(0.975, 0, 50, a = 0))
```


To sum up, we are going to use the following priors:

 \begin{equation}
 \begin{aligned}
 \alpha &\sim \mathit{Normal}(0,10)\\
 \beta  &\sim \mathit{Normal}(0,10)\\
 \sigma  &\sim \mathit{Normal}_{+}(0,50)
 \end{aligned}
 \end{equation}


A model such as $M_{cp}$ is sometimes called a \index{Fixed-effects model} *fixed-effects* model: all the parameters are fixed in the sense that do not vary from subject to subject or from item to item. A similar frequentist model  would correspond to fitting a simple linear model using the `lm()` function: `lm(n400 ~ 1 + c_cloze, data = df_eeg)`.

We fit this model in \index{brms} `brms` as follows (the default family is `gaussian()` so we can omit it). As with the `lm()` function in `R`, by default an intercept is fitted and thus `n400 ~ c_cloze` is equivalent to `n400 ~ 1 + c_cloze`:

```{r, message = FALSE, results = "hide"}
fit_N400_cp <-
  brm(n400 ~ c_cloze,
      prior = c(prior(normal(0, 10), class = Intercept),
                prior(normal(0, 10), class = b, coef = c_cloze),
                prior(normal(0, 50), class = sigma)),
      data = df_eeg)
```


For now, check the summary, and plot the posteriors of the model (Figure \@ref(fig:figurefitN400cp)).

```{r, eval = FALSE}
fit_N400_cp
```

```{r, echo = FALSE}
short_summary(fit_N400_cp)
```

(ref:figurefitN400cp) Posterior distributions of the complete pooling model, `fit_N400_cp`.

```{r figurefitN400cp, fig.cap="(ref:figurefitN400cp)",fig.height = 3.5}
plot(fit_N400_cp)
```

### \index{No pooling model} No pooling model ($M_{np}$)

One of the assumptions of the previous model is clearly wrong: observations are not independent, they are clustered by subject (and also by the specific item, but we'll ignore this until section \@ref(sec-mcvivs)). It is reasonable to assume that EEG signals are more similar within subjects than between them. The following model assumes that each subject is completely independent from each other.^[For simplicity, we assume that they share the same standard deviation.]

#### Model assumptions

1. EEG averages for the N400 spatio-temporal window are normally distributed.
2. Every subject's model is fit independently of the other subjects; the subjects have no parameters in common (an exception is the standard deviation, $\sigma$; this is the same for all subjects in Equation \@ref(eq:nplik)).
3. There is a linear relationship between cloze and the EEG signal for the trial.

What likelihood and priors can we choose here?

#### Likelihood and priors

The likelihood is a normal distribution as before:

 \begin{equation}
 signal_n \sim \mathit{Normal}( \alpha_{subj[n]} + c\_cloze_n \cdot \beta_{subj[n]},\sigma)
(\#eq:nplik)
 \end{equation}

As before, $n$ represents each observation, that is, the $n$th row in the data frame, which has $N$ rows, and now the index $i$ identifies the subject. The notation  $subj[n]$, which roughly follows @GelmanHill2007, identifies the subject index; for example, if $subj[10]=3$, then the $10$th row of the data frame is from subject $3$.


We define the priors as follows:


 \begin{equation}
 \begin{aligned}
 \alpha_{i} &\sim \mathit{Normal}(0,10)\\
 \beta_{i}  &\sim \mathit{Normal}(0,10)\\
 \sigma  &\sim \mathit{Normal}_+(0,50)
 \end{aligned}
 \end{equation}

In `brms`, such a model can be fit by removing the common intercept with the formula `n400 ~ 0 + factor(subj) + c_cloze:factor(subj)`.

This formula forces the model to estimate  one intercept and one slope for *each* level of `subj`. ^[If we don't remove the intercept, that is, if we use the formula `n400 ~ 1 + factor(subj) + c_cloze:factor(subj)`, with `factor(subj)` we are going estimate the deviation between the first subject and each of the other subjects. This is because the default contrast coding for the subjects, treated as a fixed effect, is treatment contrasts. See chapters \@ref(ch-contr) and  \@ref(ch-coding2x2).]
The by-subject intercepts are indicated with `factor(subj)` and the by-subject slopes with `c_cloze:factor(subj)`. 

It's very important to specify that `subject` should be treated as a factor and not as a numerical variable; we don't assume that subject number 3 will show 3 times more positive (or negative) average signal than subject number 1! The model fits 37 independent intercepts and 37 independent slopes. By setting a prior to  `class = b` and omitting `coef`, we are essentially setting identical priors to all the intercepts and slopes of the model. The parameters are  independent from each other; it is only our previous knowledge (or prior beliefs) about their possible values (encoded in the priors) that is identical. We can set different priors to each intercept and slope, but that will mean setting `r 37*2` priors!


```{r, message = FALSE, results = "hide"}
fit_N400_np <- brm(n400 ~ 0 + factor(subj) + c_cloze:factor(subj),
                   prior = c(prior(normal(0, 10), class = b),
                             prior(normal(0, 50), class = sigma)),
                   data = df_eeg)
```

For this model, printing a summary means printing the `r 37*2 +1` parameters ($\alpha_{1,...,37}$, $\beta_{1,...,37}$, and $\sigma$). We could do this as always by printing out the model results: just type `fit_N400_np`.

It may be easier to understand the output of the model by plotting $\beta_{1,..,37}$ using `bayesplot`. (`brms` also  includes a wrapper for this function called `stanplot`.) We can take a look at the internal names that `brms` gives to the parameters with `variables(fit_N400_np)`; they are `b_factorsubj`, then the subject index and then `:c_cloze`.  The code below changes the subject labels back to their original numerical indices and plots them in Figure \@ref(fig:nopooling). The subjects are ordered by the magnitude of their mean effects.

The model $M_{np}$ does not estimate a unique population-level effect; instead, there is a different effect estimated for each subject. However, given the posterior means from each subject, it is still possible to calculate the average of these estimates $\hat\beta_{1,...,I}$, where $I$ is the total number of subjects:



```{r}
# parameter name of beta by subject:
ind_effects_np <- paste0("b_factorsubj",
                         unique(df_eeg$subj), ":c_cloze")
beta_across_subj <- as.data.frame(fit_N400_np) %>%
  #removes the meta data from the object
  select(all_of(ind_effects_np)) %>%
  rowMeans()

# Calculate the average of these estimates
(grand_av_beta <- tibble(mean = mean(beta_across_subj),
                         lq = quantile(beta_across_subj, c(0.025)),
                         hq = quantile(beta_across_subj, c(0.975))))
```

In Figure \@ref(fig:nopooling), the 95% credible interval of this overall mean effect is plotted as two vertical lines together with the effect of cloze probability for each subject (ordered by effect size). Here, rather than using a plotting function from `brms`, we can extract the summary of by-subject effects, reorder them by magnitude, and then plot the summary with a custom plot using `ggplot2`.

```{r message = FALSE}
# make a table of beta's by subject
beta_by_subj <- posterior_summary(fit_N400_np,
                                  variable = ind_effects_np) %>%
  as.data.frame() %>%
  mutate(subject = 1:n()) %>%
  ## reorder plot by magnitude of mean:
  arrange(Estimate) %>%
  mutate(subject = factor(subject, levels = subject))
```

The code below generates Figure \@ref(fig:nopooling).

(ref:nopooling) 95% credible intervals of the effect of cloze probability for each subject according to the no pooling model, `fit_N400_np`. The solid vertical line represents the mean over all the subjects; and the broken vertical lines mark the 95% credible interval for this mean.

```{r nopooling,fig.height=9.5, message=FALSE, fig.cap="(ref:nopooling)"}
ggplot(beta_by_subj,
       aes(x = Estimate, xmin = Q2.5, xmax = Q97.5, y = subject)) +
  geom_point() +
  geom_errorbarh() +
  geom_vline(xintercept = grand_av_beta$mean) +
  geom_vline(xintercept = grand_av_beta$lq, linetype = "dashed") +
  geom_vline(xintercept = grand_av_beta$hq, linetype = "dashed") +
  xlab("By-subject effect of cloze probability in microvolts")
```

### \index{Varying intercepts} Varying intercepts and \index{Varying slopes} varying slopes model ($M_{v}$) {#sec-uncorrelated}


One major problem with the no-pooling model is that we fit each subject's data ignoring the information available in the other subjects' data.
Depending on the characteristics of the data set, the Bayesian no-pooling model might either underfit or overfit the individual subjects' data. It will ignore what can be learned from all the data taken together, and might end up in one of these two situations: if there is not enough data for some subjects, it will underfit—relying too much on the priors and learning too little from the data. Alternatively, if there is enough data, it might overfit, that is, it might learn "too much" from the individual subjects (the estimates from some subjects might be quite extreme and unrepresentative), leading us to overinterpret noisy estimates from the subjects' data [Type M error; see @gelmancarlin]. The model can be modified to explicitly assume that the subjects have an overall effect common to all the subjects, with the individual subjects deviating from this common effect.

In the model that we fit next, we will assume that there is an overall effect that is common to the subjects and, importantly, that all subjects' parameters originate from one common (normal) distribution. This model specification will result in the estimation of posteriors for each subject being also influenced by what we know about all the subjects together. We begin with a hierarchical model with uncorrelated  varying intercepts and slopes. The analogous frequentist model can be fit using `lmer()` from the package `lme4`, using `(1 + c_cloze || subj)` or, equivalently, `(c_cloze || subj)` for the by-subject random effects.

#### Model assumptions

1. EEG averages for the N400 spatio-temporal window are normally distributed.
2. Each subject deviates to some extent (this is made precise below) from the grand mean and from the mean effect of predictability. This implies that there is some between-subject variability in the individual-level intercept and slope adjustments by subject.
3. There is a linear relationship between cloze and the EEG signal.

#### Likelihood and priors

The likelihood now incorporates the assumption that both the intercept and slope are adjusted by subject.

 \begin{equation}
  signal_n \sim \mathit{Normal}(\alpha + u_{subj[n],1} + c\_cloze_n \cdot (\beta+ u_{subj[n],2}),\sigma)
 \end{equation}

\begin{equation}
 \begin{aligned}
 \alpha &\sim \mathit{Normal}(0,10)\\
 \beta  &\sim \mathit{Normal}(0,10)\\
 u_1 &\sim \mathit{Normal}(0,\tau_{u_1})\\
 u_2 &\sim \mathit{Normal}(0,\tau_{u_2})\\
 \tau_{u_1} &\sim \mathit{Normal}_+(0,20) \\
 \tau_{u_2} &\sim \mathit{Normal}_+(0,20) \\
 \sigma  &\sim \mathit{Normal}_+(0,50)
 \end{aligned}
 \end{equation}

In this model each subject has their own intercept adjustment, $u_{subj,1}$, and slope adjustment, $u_{subj,2}$.^[The intercept adjustment is often called $u_0$ in statistics books, where the intercept might be called $\alpha$ (or sometimes also $\beta_0$), and thus $u_1$ refers to the adjustment to the slope. However, in this book, we start the indexing with 1 to be consistent with the Stan language.] If $u_{subj,1}$ is positive, the subject will have a more positive EEG signal than the grand mean average. If $u_{subj,2}$ is positive, the subject will have a more positive EEG response to a change of one unit in `c_cloze`  than the overall mean effect (i.e., there will be a more positive effect of cloze probability on the N400). The parameters $u$ are sometimes called \index{Random effects} random effects and thus a model with \index{Fixed effects} fixed effects ($\alpha$ and $\beta$) and random effects is called a \index{Mixed model} mixed model. However, random effects have different meanings in different contexts. To avoid ambiguity, `brms` calls these random-effects parameters \index{Group-level effects} *group-level* effects. Since we are estimating $\alpha$ and $u$ at the same time and we assume that the average of the $u$'s is $0$ (since it is assumed to be normally distributed with mean $0$), what is common between the subjects, the grand mean, is estimated as the intercept $\alpha$, and the deviations of individual subjects' means from this grand mean are the adjustments $u_1$. Similarly, the mean effect of cloze is estimated as $\beta$, and the deviations of individual subjects' mean effects of cloze from $\beta$ are the adjustment $u_2$. The standard deviations of these two adjustment terms, $\tau_{u_1}$ and $\tau_{u_2}$, respectively, represent between subject variability.

Thus, the model $M_{v}$ has three *standard deviations*: $\sigma$, $\tau_{u_1}$ and $\tau_{u_2}$. In statistics, it is conventional to talk about variances (the square of these standard deviations); for this reason, these standard deviations are also (confusingly) called \index{Variance component} *variance components*. The variance components $\tau_{u_1}$ and $\tau_{u_2}$ characterize \index{Between-subject variability} between-subject variability, and the variance component $\sigma$ characterizes \index{Within-subject variability} within-subject variability.

The by-subject adjustments $u_1$ and $u_2$ are parameters in the model, and therefore have priors defined on them. By contrast, in the frequentist `lmer` model, the adjustments $u_1$ and $u_2$ are not parameters; they are called conditional modes; see @R-lme4.

Parameters that appear in the prior specifications for parameters, such as $\tau_u$, are often called \index{Hyperparameter} *hyperparameters*,^[Another potential source of confusion here is that *hyperparameters* is also used in the machine learning literature with a different meaning.] and the  priors on such hyperparameters are called \index{Hyperprior} *hyperpriors*. Thus, the parameter $u_1$ has $\mathit{Normal}(0,\tau_{u_1})$ as a prior; $\tau_{u_1}$ is a hyperparameter, and the hyperprior on $\tau_{u_1}$ is $\mathit{Normal}(0,20)$.^[One could in theory keep going deeper and deeper, defining hyper-hyperpriors etc., but the model would quickly become impossible to fit.]

We know that in general, in EEG experiments, the standard deviations for the by-subject adjustments are smaller than the standard deviation of the observations (which is the within-subjects standard deviation). That is, usually the between-subject variability in the intercepts and slopes is smaller than the within-subjects variability in the data. For this reason, reducing the scale of the truncated normal distribution to $20$ (in comparison to $50$) seems reasonable for the priors of the $\tau$ parameters. As always, we can do a sensitivity analysis to verify that our priors are reasonably uninformative (if we intended them to be uninformative).


For now, we are assuming that there is no relationship (no correlation) between the by-subject intercept and slope adjustments $u_1$ and $u_2$;  this lack of  correlation is indicated in `brms` using the double pipe `||`. The double pipe is also used in the same way in `lmer()` from the package `lme4` (in fact `brms` bases its syntax on that of the `lme4` package).

In \index{brms} `brms`, we need to specify hyperpriors for $\tau_{u_1}$ and $\tau_{u_2}$; these are called `sd` in `brms`, to distinguish these standard deviations from the standard deviation of the residuals $\sigma$. As with the population-level effects, the by-subjects intercept adjustments are implicitly fit for the group-level effects and thus `(c_cloze || subj)` is equivalent to `(1 + c_cloze || subj)`. If we don't want an intercept we need to explicitly indicate this with `(0 + c_cloze || subj)` or `(-1 + c_cloze || subj)`. Such a removal of the intercept is not normally done.

```{r, message = FALSE, results = "hide"}
prior_v <-
  c(prior(normal(0, 10), class = Intercept),
    prior(normal(0, 10), class = b, coef = c_cloze),
    prior(normal(0, 50), class = sigma),
    prior(normal(0, 20), class = sd, coef = Intercept, group = subj),
    prior(normal(0, 20), class = sd, coef = c_cloze, group = subj))

fit_N400_v <- brm(n400 ~ c_cloze + (c_cloze || subj),
                  prior = prior_v,
                  data = df_eeg)
```


When we print a  `brms` fit, we first  see  the summaries of the posteriors of the standard deviation of the by-group intercept and slopes,  $\tau_{u_1}$ and $\tau_{u_2}$ as `sd(Intercept)` and `sd(c_cloze)`, and then, as with previous models, the population-level effects, $\alpha$ and $\beta$ as `Intercept` and `c_cloze`, and the scale of the likelihood, $\sigma$, as `sigma`. The full summary can be printed out by typing:

```{r eval=FALSE}
fit_N400_v
```

Because the above command will result in some pages of output, it is easier to understand the summary graphically (Figure \@ref(fig:plotfitN400v)). Rather than the wrapper `plot()`, we use the original function of the package \index{\texttt{bayesplot}} `bayesplot`, \index{\texttt{mcmc\_dens}} `mcmc_dens()`, to only show density plots. We extract the first 5 parameters of the model with `variables(fit_N400_v)[1:5]`.

(ref:plotfitN400v) Posterior distributions of the parameters in the model `fit_N400_v`.

```{r plotfitN400v,fig.cap="(ref:plotfitN400v)", fig.height = 3.5}
mcmc_dens(fit_N400_v, pars = variables(fit_N400_v)[1:5])
```

Because we estimated how the population-level effect of cloze is adjusted for each subject, we could examine how each subject is being affected by the manipulation. For this we do the following, and we plot it in Figure \@ref(fig:partialpooling). These are adjustments, $u_{1,2},u_{2,2},...,u_{37,2}$, and not the effect of the manipulation by subject, $\beta + [u_{1,2},u_{2,2},...,u_{37,2}]$.  The code below produces Figure \@ref(fig:partialpooling).

(ref:partialpooling) 95% credible intervals of adjustments to the effect of cloze probability for each subject ($u_{1..37,2}$) according to the varying intercept and varying slopes model, `fit_N400_v`. To obtain the effect of cloze probability for each subject, we would need to add the estimate of $\beta$ to each adjustment.

```{r partialpooling, fig.cap = "(ref:partialpooling)", fig.height=9.5,message=FALSE,warning=FALSE}
# make a table of u_2s
ind_effects_v <- paste0("r_subj[", unique(df_eeg$subj),
                        ",c_cloze]")
u_2_v <- posterior_summary(fit_N400_v, variable = ind_effects_v) %>%
  as_tibble() %>%
  mutate(subj = 1:n()) %>%
  ## reorder plot by magnitude of mean:
  arrange(Estimate) %>%
  mutate(subj = factor(subj, levels = subj))
# We plot:
ggplot(u_2_v,
       aes(x = Estimate, xmin = Q2.5, xmax = Q97.5, y = subj)) +
  geom_point() +
  geom_errorbarh() +
  xlab("By-subject adjustment to the slope in microvolts")
```

There is an important difference between the no-pooling model and the varying intercepts and slopes model we just fit. The no-pooling model fits each individual subject's intercept and slope independently for each subject. By contrast, the varying intercepts and slopes model takes *all* the subjects' data into account in order to compute the fixed effects $\alpha$ and $\beta$; and the model "shrinks" [@pinheirobates] the by-subject intercept and slope adjustments towards the fixed effects estimates. In Figure \@ref(fig:comparison), we can see the \index{Shrinkage} shrinkage of the estimates in the varying intercepts model by comparing them with the estimates of the no pooling model ($M_{np}$).

```{r, fold = TRUE}
# Extract parameter estimates from the no pooling model:
par_np <- posterior_summary(fit_N400_np, variable = ind_effects_np) %>%
  as_tibble() %>%
  mutate(model = "No pooling",
         subj = unique(df_eeg$subj))
# For the hierarchical model, the code is more complicated
# because we want the effect (beta) + adjustment.
# Extract the overall group level effect:
beta <- c(as_draws_df(fit_N400_v)$b_c_cloze)
# Extract the individual adjustments:
ind_effects_v <- paste0("r_subj[", unique(df_eeg$subj), ",c_cloze]")
adjustment <- as_draws_matrix(fit_N400_v, variable = ind_effects_v)
# Get the by subject effects in a data frame where each adjustment
# is in each column.
# Remove all the draws meta data by using as.data.frame
by_subj_effect <- as.data.frame(beta + adjustment)
# Summarize them by getting a table with the mean and the
# quantiles for each column and then binding them.
par_h <- lapply(by_subj_effect, function(x) {
  tibble(Estimate = mean(x),
         Q2.5 = quantile(x, .025),
         Q97.5 = quantile(x, .975))}) %>%
  bind_rows() %>%
  # Add a column to identify that the model,
  # and one with the subject labels:
  mutate(model = "Hierarchical",
         subj = unique(df_eeg$subj))
# The mean and 95% CI of both models in one data frame:
by_subj_df <- bind_rows(par_h, par_np) %>%
  arrange(Estimate) %>%
  mutate(subj = factor(subj, levels = unique(.data$subj)))
```

(ref:comparison) This plot compares the estimates of the effect of cloze probability for each subject between (i) the no pooling, `fit_N400_np` and (ii) the varying intercepts and varying slopes, hierarchical, model, `fit_N400_v`.

```{r comparison, message=FALSE, fig.height=9.5, fig.cap= "(ref:comparison)"}
b_c_cloze <- posterior_summary(fit_N400_v, variable = "b_c_cloze")
ggplot(by_subj_df,
       aes(ymin = Q2.5, ymax = Q97.5, x = subj,
           y = Estimate, color = model, shape = model)) +
  geom_errorbar(position = position_dodge(1)) +
  geom_point(position = position_dodge(1)) +
  # We'll also add the mean and 95% CrI of the overall difference
  # to the plot:
  geom_hline(yintercept = b_c_cloze[, "Estimate"]) +
  geom_hline(yintercept = b_c_cloze[, "Q2.5"],
             linetype = "dotted", linewidth = .5) +
  geom_hline(yintercept = b_c_cloze[, "Q97.5"],
             linetype = "dotted", linewidth = .5) +
  xlab("N400 effect of predictability") +
  coord_flip()
```

### Correlated varying intercept varying slopes model ($M_{h}$) {#sec-mcvivs}


The model $M_{v}$ allowed for differences in intercepts (mean voltage) and slopes (effects of cloze) across subjects, but it has the implicit assumption that these varying intercepts and varying slopes are independent. It is in principle possible that subjects showing more negative voltage may also show stronger effects (or weaker effects). Next, we fit a model that allows a correlation between the intercepts and slopes. We model the \index{Correlation between varying intercepts and slopes} correlation between varying intercepts and slopes by defining a variance-covariance matrix  $\boldsymbol{\Sigma}$ between the by-subject varying intercepts and slopes, and by assuming that both adjustments (intercept and slope) come from a \index{Multivariate} multivariate (in this case, a bivariate) \index{Normal distribution} normal distribution.

* In $M_h$, we  model the EEG data with the following assumptions:

1. EEG averages for the N400 spatio-temporal window are normally distributed.
2. Some aspects of the mean signal voltage and of the effect of predictability depend on the subject, and these two might be correlated, i.e., we assume group-level intercepts and slopes, and allow a correlation between them by-subject.
3. There is a linear relationship between cloze and the EEG signal for the trial.


The likelihood remains identical to the model $M_v$, which assumes no correlation between group-level intercepts and slopes (section \@ref(sec-uncorrelated)):

 \begin{equation}
  signal_n \sim \mathit{Normal}(\alpha + u_{subj[n],1} + c\_cloze_n \cdot  (\beta + u_{subj[n],2}),\sigma)
  \end{equation}


The correlation is indicated in the  priors on the adjustments for intercept $u_{1}$ and slopes $u_{2}$.

* Priors:
 \begin{equation}
 \begin{aligned}
   \alpha & \sim \mathit{Normal}(0,10) \\
   \beta  & \sim \mathit{Normal}(0,10) \\
    \sigma  &\sim \mathit{Normal}_+(0,50)\\
    {\begin{pmatrix}
    u_{i,1} \\
    u_{i,2}
    \end{pmatrix}}
   &\sim {\mathcal {N}}
    \left(
   {\begin{pmatrix}
    0\\
    0
   \end{pmatrix}}
 ,\boldsymbol{\Sigma_u} \right)
 \end{aligned}
 \end{equation}


In this model, a \index{Bivariate normal distribution} bivariate normal distribution generates the varying intercepts and varying slopes $\mathbf{u}$; this is an $n\times 2$ matrix. The  \index{Variance-covariance matrix} variance-covariance matrix $\boldsymbol{\Sigma_u}$ defines the standard deviations of the varying intercepts and varying slopes, and the correlation between them.  Recall from section \@ref(sec-contbivar) that the diagonals of the variance-covariance matrix contain the variances of the correlated random variables, and the off-diagonals contain the \index{Covariance} covariances. In this example, the covariance $Cov(u_1,u_2)$ between two variables $u_1$ and $u_2$ is defined as the product of their \index{Correlation} correlation $\rho$ and their standard deviations $\tau_{u_1}$ and $\tau_{u_2}$. In other words, $Cov(u_1,u_2) = \rho_u \tau_{u_1} \tau_{u_2}$.

\begin{equation}
\boldsymbol{\Sigma_u} =
{\begin{pmatrix}
\tau_{u_1}^2 & \rho_u \tau_{u_1} \tau_{u_2} \\
\rho_u \tau_{u_1} \tau_{u_2} & \tau_{u_2}^2
\end{pmatrix}}
\end{equation}


In order to specify a prior for $\Sigma_u$, we need priors for the standard deviations, $\tau_{u_1}$, and $\tau_{u_2}$, and also for their correlation, $\rho_u$. We can use the same priors for $\tau$ as before. For the correlation matrix that contains $\rho_u$, we use the  \index{LKJ prior} LKJ prior. The basic idea of the LKJ prior on the correlation matrix is that as its parameter, $\eta$ (*eta*), increases, it will favor correlations closer to zero.^[This is because an LKJ correlation distribution with a large $\eta$ corresponds to a correlation matrix with values close to zero in the lower and upper triangles.] At $\eta = 1$, the LKJ correlation distribution is uninformative (if there is a single correlation parameter, this is similar to $\mathit{Beta}(1,1)$, scaled to the interval $[-1,1]$), at $\eta < 1$, it favors extreme correlations  (similar to $\mathit{Beta}(a<1,b<1)$). We set $\eta = 2$ so that we don't favor extreme correlations, and we still represent our lack of knowledge through the wide spread of the prior between $-1$ and $1$. Thus, $\eta = 2$ gives us a regularizing, relatively uninformative or mildly informative prior.

Figure \@ref(fig:lkjviz) shows a visualization of different parameterizations of the LKJ prior.

(ref:lkjviz) Visualization of the LKJ correlation distribution prior with four different values of the $\eta$ parameter.

```{r lkjviz,echo=FALSE, fig.cap ="(ref:lkjviz)", message= FALSE,warning=FALSE,results="asis",cache=TRUE, fig.width =3, fig.height =3,fig.show='hold', out.width='48%'}
## https://github.com/rmcelreath/rethinking/blob/1def057174071beb212532d545bc2d8c559760a2/R/distributions.r
# onion method correlation matrix
dlkjcorr <- function(x, eta = 1, log = FALSE) {
  ll <- det(x)^(eta - 1)
  if (log == FALSE) ll <- exp(ll)
  return(ll)
}

# Simplified for a 2 x 2 matrix
dlkjcorr2 <- function(rho, eta = 1) {
  map_dbl(rho, ~ matrix(c(1, .x, .x, 1), ncol = 2) %>%
    dlkjcorr(., eta))
}

ggplot(tibble(rho = c(-0.99, 0.99)), aes(rho)) +
  stat_function(fun = dlkjcorr2, geom = "line", args = list(eta = 1)) +
  ylab("density") +
  ggtitle("eta = 1")

ggplot(tibble(rho = c(-0.99, 0.99)), aes(rho)) +
  stat_function(fun = dlkjcorr2, geom = "line", args = list(eta = 2)) +
  ylab("density") +
  ggtitle("eta = 2")

ggplot(tibble(rho = c(-0.99, 0.99)), aes(rho)) +
  stat_function(fun = dlkjcorr2, geom = "line", args = list(eta = 4)) +
  ylab("density") +
  ggtitle("eta = 4")

ggplot(tibble(rho = c(-0.99, 0.99)), aes(rho)) +
  stat_function(fun = dlkjcorr2, geom = "line", args = list(eta = .9)) +
  ylab("density") +
  ggtitle("eta = .9")
```

\begin{equation}
\begin{aligned}
\tau_{u_1} &\sim \mathit{Normal}_+(0,20)\\
\tau_{u_2} &\sim \mathit{Normal}_+(0,20)\\
\begin{bmatrix}
1 & \rho_u \\
\rho_u & 1
\end{bmatrix} &\sim \mathit{LKJcorr}(2)
\end{aligned}
\end{equation}



In our \index{brms} `brms` model, we allow a correlation  between the by-subject intercepts and slopes  by using a single pipe `|` instead of the double pipe `||` that we used previously. This convention follows the syntax used in the frequentist `lmer()` function. As before, the varying intercepts are implicitly fit.

Because we have a new parameter, the correlation  $\rho_{u}$, we need to add a new prior for this correlation. In `brms`, this is achieved by adding a prior for the parameter type `cor`. This is not obvious from the `brms` syntax, but the LKJ prior is being defined on the correlation matrix, not the correlation $\rho_{u}$.

```{r, message = FALSE, results = "hide"}
prior_h <- c(prior(normal(0, 10), class = Intercept),
             prior(normal(0, 10), class = b, coef = c_cloze),
             prior(normal(0, 50), class = sigma),
             prior(normal(0, 20),
                   class = sd, coef = Intercept,
                   group = subj),
             prior(normal(0, 20),
                   class = sd, coef = c_cloze,
                   group = subj),
             prior(lkj(2), class = cor, group = subj))

fit_N400_h <- brm(n400 ~ c_cloze + (c_cloze | subj),
                  prior = prior_h,
                  data = df_eeg)
```

The estimates do not change much in comparison with the varying intercept and varying slope model, probably because the estimation of the correlation is quite poor (i.e., there is a lot of uncertainty in the posterior). While the inclusion of the correlation parameter is justified, this results, in our experience, in a slower convergence of the models (see also the discussion at the end of the section \@ref(sec-distrmodel)).
As before, the estimates are shown graphically, in Figure \@ref(fig:plotfitN400h). As always, one can access the complete summary by typing `fit_N400_h`.

(ref:plotfitN400h) The posteriors of the parameters in the model `fit_N400_h`.

```{r plotfitN400h,fig.cap="(ref:plotfitN400h)", fig.height = 6.5}
plot(fit_N400_h, nvariables = 6)
```

We are now half-way to what is sometimes called the \index{Maximal hierarchical model} "maximal" hierarchical model [@barr2013]. This usually refers to a model with all the by-participant and by-items group-level variance components *allowed by the experimental design* and *a full variance covariance matrix* for all the group-level parameters. Not all variance components are allowed by the experimental design: in particular, between-group manipulations cannot have variance components. For example, even if we assume that the working memory capacity of the subjects might affect the N400, we cannot measure how working memory affects the subjects differently.

When we refer to a full variance-covariance matrix, we mean a variance-covariance matrix where all the elements (variances and covariances) are non-zero. In our previous model, for example, the variance-covariance matrix $\boldsymbol{\Sigma_u}$ was full because no element was zero. If we assume no correlation between group-level intercept and slope, it would mean to have zeros in the diagonal of the matrix and this would render the model to be identical to $M_{v}$ defined in section \@ref(sec-uncorrelated); if we assume that also the bottom right element ($\tau^2$) is zero, the model would turn into a \index{Varying intercept model} varying intercept model (in `brms` formula `n400 ~ c_cloze + (1 | subj)`); and if we assume that the matrix has only zeros, the model would turn into a complete pooling model, $M_{cp}$, as defined in section \@ref(sec-Mcp).

As we will see in section \@ref(sec-distrmodel) and in the online chapter \@ref(ch-workflow), "maximal" is probably a misnomer for Bayesian models, since this mostly refers to limitations of the popular frequentist package for fitting models, `lme4`.

The next section spells out a model with full variance-covariance matrix for both subjects and items-level effects.

### By-subjects and by-items correlated varying intercept varying slopes model ($M_{sih}$) {#sec-sih}

Our new model, $M_{sih}$ will allow for differences in intercepts (mean voltage) and slopes (effects of predictability) across subjects and across items. In typical Latin square designs, subjects and items are said to be \index{Crossed random effects} *crossed random effects*---each subject sees exactly one instance of each item. Here  we assume a possible \index{Correlation} correlation between varying intercepts and slopes by subjects, and another one by items.


* In $M_{sih}$, we model the EEG data with the following assumptions:


1. EEG averages for the N400 spatio-temporal window are normally distributed.
2. Some aspects of the mean signal voltage and of the effect of predictability depend on the subject, i.e., we assume group-level intercepts, and slopes, and a correlation between them by-subject.
2. Some aspects of the mean signal voltage and of the effect of predictability depend on the item, i.e., we assume group-level intercepts, and slopes, and a correlation between them by-item.

4. There is a linear relationship between cloze and the EEG signal for the trial.


* Likelihood:

 \begin{multline}
  signal_n \sim \mathit{Normal}(\alpha + u_{subj[n],1} + w_{item[n],1} + \\ c\_cloze_n \cdot  (\beta + u_{subj[n],2}+ w_{item[n],2}), \sigma)
  \end{multline}



* Priors:
 \begin{equation}
 \begin{aligned}
   \alpha & \sim \mathit{Normal}(0,10) \\
   \beta  & \sim \mathit{Normal}(0,10) \\
    \sigma  &\sim \mathit{Normal}_+(0,50)\\
    {\begin{pmatrix}
    u_{i,1} \\
    u_{i,2}
    \end{pmatrix}}
   &\sim {\mathcal {N}}
    \left(
   {\begin{pmatrix}
    0\\
    0
   \end{pmatrix}}
 ,\boldsymbol{\Sigma_u} \right) \\
     {\begin{pmatrix}
    w_{j,1} \\
    w_{j,2}
    \end{pmatrix}}
   &\sim {\mathcal {N}}
    \left(
   {\begin{pmatrix}
    0\\
    0
   \end{pmatrix}}
 ,\boldsymbol{\Sigma_w} \right)
 \end{aligned}
 \end{equation}

We have added the index  $j$, which represents each item, as we did with subjects;  $item[n]$ indicates the item that corresponds to the observation in the $n$-th row of the data frame.

We have hyperparameters and hyperpriors as before:

\begin{equation}
\begin{aligned}
 \boldsymbol{\Sigma_u} & =
{\begin{pmatrix}
\tau_{u_1}^2 & \rho_u \tau_{u_1} \tau_{u_2} \\
\rho_u \tau_{u_1} \tau_{u_2} & \tau_{u_2}^2
\end{pmatrix}}\\
 \boldsymbol{\Sigma_w} & =
{\begin{pmatrix}
\tau_{w_1}^2 & \rho_w \tau_{w_1} \tau_{w_2} \\
\rho_w \tau_{w_1} \tau_{w_2} & \tau_{w_2}^2
\end{pmatrix}}
 \end{aligned}
\end{equation}

\begin{equation}
\begin{aligned}
\tau_{u_1} &\sim \mathit{Normal}_+(0,20)\\
\tau_{u_2} &\sim \mathit{Normal}_+(0,20)\\
\begin{bmatrix}
1 & \rho_u \\
\rho_u & 1
\end{bmatrix} \sim \mathit{LKJcorr}(2)\\
\tau_{w_1} &\sim \mathit{Normal}_+(0,20)\\
\tau_{w_2} &\sim \mathit{Normal}_+(0,20)\\
\begin{bmatrix}
1 & \rho_w \\
\rho_w & 1
\end{bmatrix} \sim \mathit{LKJcorr}(2)\\
\end{aligned}
\end{equation}

We set identical priors for by-items group-level effects as for  by-subject group-level effects, but this only because we don't have any differentiated prior information about subject-level vs. item-level variation. However, bear in mind that the estimation for items is completely independent from the estimation for subjects. Although we wrote many more equations than before, the `brms` model is quite straightforward to extend. 

We assign the priors to the by-subject and by-item parameters first:

```{r, eval = FALSE}
prior_sih_full <-
  c(prior(normal(0, 10), class = Intercept),
    prior(normal(0, 10), class = b, coef = c_cloze),
    prior(normal(0, 50), class = sigma),
    prior(normal(0, 20),
          class = sd, coef = Intercept,
          group = subj),
    prior(normal(0, 20),
          class = sd, coef = c_cloze,
          group = subj),
    prior(lkj(2), class = cor, group = subj),
    prior(normal(0, 20),
          class = sd, coef = Intercept,
          group = item),
    prior(normal(0, 20),
          class = sd, coef = c_cloze,
          group = item),
    prior(lkj(2), class = cor, group = item))
```

We can also simplify the priors, when we assign the same priors to the by-subject and by-item parameters:

```{r}
prior_sih <-
  c(prior(normal(0, 10), class = Intercept),
    prior(normal(0, 10), class = b),
    prior(normal(0, 50), class = sigma),
    prior(normal(0, 20), class = sd),
    prior(lkj(2), class = cor))
```

Fit the model:

```{r, message = FALSE, results = "hide", eval = !file.exists("dataR/fit_N400_sih.RDS")}
fit_N400_sih <- brm(n400 ~ c_cloze + (c_cloze | subj) +
                      (c_cloze | item),
                    prior = prior_sih,
                    data = df_eeg)
```

```{r, echo= FALSE, eval = TRUE}
if(!file.exists("dataR/fit_N400_sih.RDS")){
  saveRDS(fit_N400_sih, "dataR/fit_N400_sih.RDS")
} else {
  fit_N400_sih <- readRDS("dataR/fit_N400_sih.RDS")
}
```


We have new group-level effects in the summary if we run `fit_N400_sih`, but again the estimate of the effect of cloze remains virtually unchanged (Figure \@ref(fig:fitN400sih)).

```{r, eval = FALSE}
fit_N400_sih
```

```{r, echo = FALSE}
short_summary(fit_N400_sih)
```


(ref:fitN400sih) The posterior distributions of the parameters in the model `fit_N400_sih`.

```{r fitN400sih, fig.height = 11,fig.cap="(ref:fitN400sih)"}
plot(fit_N400_sih, nvariables = 9)
```

### Beyond the maximal model--Distributional regression models {#sec-distrmodel}

\index{Distributional regression}

We can use posterior predictive checks to verify that our last model can capture the entire signal distribution. This is shown in Figure \@ref(fig:ppcheckdens).

(ref:ppcheckdens) Overlay of densities from the posterior predictive distributions of the model `fit_N400_sih`.

```{r ppcheckdens, fig.cap ="(ref:ppcheckdens)", fig.height = 2.2}
pp_check(fit_N400_sih, ndraws = 50, type = "dens_overlay")
```

However, we know that in ERP studies, large levels of impedance between the recording electrodes and the skin tissue increase the noise in the recordings [@picton_etal_2000]. Given that skin tissue is different between subjects, it could be the case that the level of noise varies by subject.
It might be a good idea to verify that our model is good enough for capturing the by-subject variability. The code below produces Figure \@ref(fig:postpreddensbysubj).

(ref:postpreddensbysubj) The plot shows 100 predicted distributions with the label $y_{rep}$ and the distribution of the average signal data with the label $y$ density plots for the 37 subjects that participated in the experiment.


```{r postpreddensbysubj, fig.cap ="(ref:postpreddensbysubj)" , message= FALSE,fig.height=10, echo = TRUE}
pp_check(fit_N400_sih,
          type = "dens_overlay_grouped",
          ndraws = 100,
          group = "subj") +
  xlab("Signal in the N400 spatiotemporal window") +
  theme(legend.position = "inside",
        legend.position.inside = c(.5,.08))
```

Figure \@ref(fig:postpreddensbysubj) hints that we might be misfitting some subjects: Some of the by-subject observed distributions of the EEG signal averages look much tighter than their corresponding posterior predictive distributions (e.g., subjects 3, 5, 9, 10, 14), whereas some other by-subject observed distributions look wider (e.g., subjects 25, 26, 27).  Another approach to examine whether we misfit the by-subject noise level is to  plot posterior distributions of the standard deviations and compare them with the observed standard deviation. This is achieved in the following code, which groups the data by subject, and shows the distribution of standard deviations. The result is shown in Figure  \@ref(fig:postpredsumbysubj). It is clear now that, for some subjects, the observed standard deviation lies outside the distribution of predictive standard deviations.


(ref:postpredsumbysubj) Distribution of posterior predicted standard deviations in gray and observed standard deviation in black lines by subject.

```{r postpredsumbysubj, fig.cap ="(ref:postpredsumbysubj)", message= FALSE, fig.height=10  }
pp_check(fit_N400_sih,
         type = "stat_grouped",
         ndraws = 1000,
         group = "subj",
         stat = "sd",
         facet_args = list(scales = "fixed")) +
  scale_x_continuous(breaks = c(8, 12, 16), limits = c(7,19)) +
  theme(legend.position = "inside",
        legend.position.inside = c(.5,.08))
```

Why is our \index{Maximal hierarchical model} "maximal" hierarchical model misfitting the by-subject distribution of data? This is because, the maximal models are, in general and implicitly, models with the maximal group-level effect structure for the location parameter (e.g., the mean, $\mu$, in a normal model). Other parameters (e.g., scale or shape parameters) are estimated as auxiliary parameters, and are assumed to be constant across observations and clusters. This assumption is so common that researchers may not be aware that it is just an assumption. In the Bayesian framework, it is easy to change such default assumptions if necessary. Changing the assumption that all subjects have the same residual standard deviation leads to the  distributional regression model. Such models can be fit in `brms`; see also the `brms` vignette, https://cran.r-project.org/web/packages/brms/vignettes/brms_distreg.html.

We are going to change our previous likelihood, so that the scale, $\sigma$ has also a group-level effect structure. We exponentiate $\sigma$ to make sure that the negative adjustments do not cause $\sigma$ to become negative.

\begin{equation}
\begin{aligned}
  signal_n &\sim \mathit{Normal}(\alpha + u_{subj[n],1} + w_{item[n],1} + \\             &  \hspace{2cm} c\_cloze_n \cdot  (\beta + u_{subj[n],2}+ w_{item[n],2}), \sigma_n)\\
  \sigma_n &= \exp(\sigma_\alpha + \sigma_{u_{subj[n]}})
\end{aligned}
\end{equation}

We just need to add priors to our new parameters (that replace the old prior for $\sigma$). We set the prior to the intercept of the standard deviation, $\sigma_\alpha$, to be similar to our previous $\sigma$.  For the variance component of $\sigma$, $\tau_{\sigma_u}$, we set rather uninformative hyperpriors. Recall that everything is exponentiated when it goes inside the likelihood; that is why we use $\log(50)$ rather than $50$ for the standard deviation of the prior for  $\sigma$.

\begin{equation}
\begin{aligned}
  \sigma_\alpha &\sim \mathit{Normal}(0,\log(50))\\
  \sigma_u &\sim \mathit{Normal}(0, \tau_{\sigma_u}) \\
  \tau_{\sigma_u} &\sim \mathit{Normal}_+(0, 5)
\end{aligned}
\end{equation}

This model can be fit in `brms` using the internal function `brmsformula()` (or its shorter alias `bf()`). This is a powerful function that extends the formulas that we used so far allowing for setting a hierarchical regression to any parameter of a model. This will allow us to set a by-subject hierarchical structure to the parameter $\sigma$. We also need to set new priors; these priors are identified by `dpar = sigma`.


```{r, message = FALSE, results = "hide"}
prior_s <-
  c(prior(normal(0, 10), class = Intercept),
    prior(normal(0, 10), class = b),
    prior(normal(0, 20), class = sd),
    prior(lkj(2), class = cor),
    prior(normal(0, log(50)), class = Intercept, dpar = sigma),
    prior(normal(0, 5),
          class = sd,
          group = subj,
          dpar = sigma))

fit_N400_s <- brm(
  brmsformula(n400 ~ c_cloze + (c_cloze | subj) + (c_cloze | item),
              sigma ~ 1 + (1 | subj)),
  prior = prior_s, data = df_eeg)
```

Inspect the output below; notice that our estimate for the effect of cloze
remains very similar to that of the model `fit_N400_sih`.

Compare the two models' estimates:

```{r}
posterior_summary(fit_N400_sih, variable = "b_c_cloze")
```

```{r}
posterior_summary(fit_N400_s, variable = "b_c_cloze")
```

Nonetheless, Figure \@ref(fig:postpreddensbysubj2) shows that the fit of the model with respect to the by-subject variability is much better than before. Furthermore, Figure \@ref(fig:postpredsumbysubj2) shows that the observed standard deviations for each subject are well inside the posterior predictive distributions. The code below produces Figure \@ref(fig:postpreddensbysubj2).

(ref:postpreddensbysubj2) The gray density plots show 100 predicted distributions from a model that includes a hierarchical structure for $\sigma$. The black density plots show the distribution of the average signal data for the 37 subjects in the experiment.

```{r postpreddensbysubj2, fig.cap ="(ref:postpreddensbysubj2)" , message= FALSE, fig.height=10 }
pp_check(fit_N400_s,
          type = "dens_overlay_grouped",
          ndraws = 100,
          group = "subj") +
  xlab("Signal in the N400 spatiotemporal window") +
 theme(legend.position = "inside",
        legend.position.inside = c(.5,.08))
```

(ref:postpredsumbysubj2) The gray lines show the distributions of posterior predicted standard deviations from a model that includes a hierarchical structure for $\sigma$, and observed mean standard deviations by subject (black vertical lines).

```{r postpredsumbysubj2, fig.cap ="(ref:postpredsumbysubj2)", message= FALSE, fig.height=10, echo = FALSE }
pp_check(fit_N400_s,
         type = "stat_grouped",
         ndraws = 1000,
         group = "subj",
         stat = "sd",
         facet_args = list(scales = "fixed")) +
  scale_x_continuous(breaks = c(8, 12, 16), limits = c(7,19)) +
  theme(legend.position = "inside",
        legend.position.inside = c(.5,.08))
```

The model `fit_N400_s` raises the question: how much structure should we add to our statistical model? Should we also assume that $\sigma$ can vary by items, and also by our experimental manipulation? Should we also have a maximal model for $\sigma$? Unfortunately, there are no clear answers that apply to every situation. The amount of complexity that we can introduce in a statistical model depends on (i) the answers we are looking for (we should include parameters that represent what we want to estimate), (ii) the size of the data at hand (more complex models require more data; this could mean more observations within subjects and/or the number of subjects), (iii) our computing power (as the complexity increases models take increasingly long to converge and require more computer power to finish the computations in a feasible time frame), and (iv) our domain knowledge.

Whether certain effects should be included in a model also depends on whether they are known to impact posterior inference or statistical testing (e.g., via Bayes factors). For example, it is well known that estimating group-level effects for the location parameter can have a strong influence on the test statistics for the corresponding population-level effect [@barr2013; @SchadEtAlBF]. Given that population-level effects are often what researchers care about, it is therefore important to consider group-level effects for the location parameter. However, to our knowledge, it is not clear whether estimating group-level effects for the standard deviation of the likelihood has an impact on inferences for the population-level (or fixed) effects. Maybe there is one, but it is not widely known--statistical research would have to be conducted via simulations to assess whether such an influence can take place. The point here is that for some parameters, it's crucial to include them in the model, because they are known to affect the inferences that we want to draw from the data. Other model components may (presumably) be less decisive. Which ones these are remains an open question for research.

Ultimately, all models are approximations (that's in the best case; often, they are plainly wrong) and we need to think carefully about which aspects of our data we have to account and which aspects we can abstract away from.

In the context of cognitive modeling, @mcclellandPlaceModelingCognitive2009 argues that models should not focus on a every single detail of the process they intend to explain. In order to understand a model, it needs to be simple enough. However, @mcclellandPlaceModelingCognitive2009 warns us that one must bear in mind that oversimplification does have an impact on what we can conclude from our analysis:  A simplification can limit the phenomena that a model addresses, or can even lead to incorrect predictions. There is a continuum between purely statistical models (e.g., a linear regression) and computational cognitive models. For example, we can define "hybrid" models such as the linear ballistic accumulator [@brownSimplestCompleteModel2008; and see @Nicenboim2018StanCon for an implementation in Stan], where a great deal of cognitive detail is sacrificed for tractability. The conclusions of @mcclellandPlaceModelingCognitive2009 apply to any type of model in cognitive science: "Simplification is essential, but it comes at a cost, and real understanding depends in part on understanding the effects of the simplification".

## A \index{Hierarchical log-normal model} hierarchical log-normal model: The \index{Stroop effect} Stroop effect {#sec-stroop}

Next we illustrate some of the issues that arise with a \index{Log-normal likelihood} log-normal likelihood in a hierarchical model. The data are from a Stroop task [@stroop1935studies; for a review, see @macleod1991half]. We will analyze a subset of the data of 3337 subjects that participated in one variant of the Stroop task; this was part of a battery of tasks run in @ManyLabs3.

For this variant of the Stroop task, subjects were presented with one word ("red," "blue," or "green") at the center of the screen, written in either red, blue, or green color. In one third of the trials, the word matched the text color ("congruent" condition); in the remaining trials, it did not match ("incongruent" condition). Subjects were instructed to focus only on the color of the text and press `1` if the color was red, `2` if it was blue, and `3` if it was green. The incongruent condition is more difficult because subjects need to identify the color when it mismatches the word. For example, they might need to respond that the color is blue when the word "green" is written in blue; this is more challenging than the congruent condition, where the word "green" is written in green. This increased difficulty in the incongruent condition is known as the Stroop effect, which is extremely robust across task variations.

This task yields two measures: the accuracy of the decision made, and the time it took to respond. For the Stroop task, accuracy is usually almost at ceiling; to simplify the model, we will ignore accuracy. For cognitive models that incorporates accuracy and response times into a model to analyze these Stroop data, see @Nicenboim2018StanCon. Additionally, chapters \@ref(ch-mixture) and \@ref(ch-lognormalrace) address response time and accuracy data.

### A correlated varying intercept varying slopes \index{Log-normal model} log-normal model

If our theory only focuses on the difference between the \index{Response time} response times for the "congruent" vs. "incongruent" condition, we can ignore the actual color presented and the word that was written. We can simply focus on whether a trial was congruent or incongruent. Define a predictor `c_cond` to represent these two conditions. For simplicity, we will also assume that all subjects share the same variance (as we saw in section \@ref(sec-distrmodel), changing this assumption leads to distributional regression models).

The above assumptions mean that we are going to fit the data with the following likelihood. The likelihood function is identical to the one that we fit in section \@ref(sec-mcvivs), except that here the location and scale are embedded in a log-normal likelihood rather than a normal likelihood. Equation \@ref(eq:stroopmodel) states that we are dealing with a hierarchical model with by-subjects  varying intercepts and varying slopes model:

\begin{equation}
  rt_n \sim \mathit{LogNormal}(\alpha + u_{subj[n],1}  + c\_cond_n \cdot  (\beta + u_{subj[n],2}), \sigma)
(\#eq:stroopmodel)
\end{equation}

In chapter \@ref(ch-contr), we will discuss the sum-contrast coding of the two conditions (`c_cond`). For now, it suffices to say that we assign a `+1` to `c_cond` for the "incongruent" condition, and a `-1` for the "congruent" condition (i.e., a sum-contrast coding). Under this contrast coding, if the posterior mean of the parameter $\beta$ turns out to be positive, that would mean that the model predicts that the incongruent condition has slower response times than the congruent one. This is because on average the location of the log-normal likelihood for each condition will be as follows.  In Equation \@ref(eq:stroopmodelmeans), $\mu_{incongruent}$ refers to the location of the incongruent condition, and $\mu_{congruent}$ to the location of the congruent condition.

\begin{equation}
\begin{aligned}
  \mu_{incongruent} &= \alpha + 1 \cdot  \beta \\
  \mu_{congruent} &= \alpha + -1 \cdot  \beta
  \end{aligned}
  (\#eq:stroopmodelmeans)
\end{equation}

We could have chosen to do the opposite contrast coding assignments: $-1$ for the incongruent condition, and $+1$ for congruent condition. In that case, if the posterior mean of the parameter $\beta$ turns out to be positive, that would mean that the incongruent condition has a *faster* response time than the congruent condition. Given that the Stroop effect is very robust, we do not expect such an outcome. In order to make the $\beta$ parameter easier to interpret, we have chosen the contrast coding where a positive sign on the mean of $\beta$ implies that the inconguent condition has slower response times.

As always, we need priors for all the parameters in our model. For the population-level parameters (or fixed effects), we use the same priors as we did when we were fitting a regression with a log-normal likelihood in section \@ref(sec-lognormal).

 \begin{equation}
 \begin{aligned}
   \alpha & \sim \mathit{Normal}(6, 1.5) \\
   \beta  & \sim \mathit{Normal}(0, 0.01) \\
    \sigma  &\sim \mathit{Normal}_+(0, 1)
 \end{aligned}
 \end{equation}

Here, $\beta$ represents, on the log scale, the change in the intercept $\alpha$ as a function of the experimental manipulation. In this model, $\beta$ will probably be larger in magnitude than for the model that examined the  difference in pressing the spacebar for two consecutive trials in section \@ref(sec-lognormal). We might need to examine the prior for $\beta$ with predictive distributions, but we will delay this for now.

In contrast to our previous models, the intercept $\alpha$ is not the grand mean of the location. This is because the conditions were not balanced in the experiment (one-third of the conditions were congruent  and two-thirds incongruent). The intercept could be interpreted here as the time (in log-scale) it takes to respond if we ignore the experimental manipulation.
Next, we turn our attention to the prior specification for the group-level parameters (or random effects). If we assume a possible correlation between by-subject intercepts and slopes, our model will have the following structure. In particular, we have to define priors for the parameters in the variance-covariance matrix $\Sigma_u$.

 \begin{equation}
 \begin{aligned}
    {\begin{pmatrix}
    u_{i,1} \\
    u_{i,2}
    \end{pmatrix}}
   &\sim {\mathcal {N}}
    \left(
   {\begin{pmatrix}
    0\\
    0
   \end{pmatrix}}
 ,\boldsymbol{\Sigma_u} \right)
 \end{aligned}
 \end{equation}

\begin{equation}
\begin{aligned}
 \boldsymbol{\Sigma_u} & =
{\begin{pmatrix}
\tau_{u_1}^2 & \rho_u \tau_{u_1} \tau_{u_2} \\
\rho_u \tau_{u_1} \tau_{u_2} & \tau_{u_2}^2
\end{pmatrix}}
\end{aligned}
\end{equation}

In practice, this means that we need priors for the by-subject standard deviations and correlations. For the variance components, we will set a similar prior as for $\sigma$. We don't expect the by-group adjustments to the intercept and slope  to have more variance than the within-subject variance, so this prior will be quite conservative because it allows for a large range of prior uncertainty. We assign the same prior for the correlations as we did in section \@ref(sec-sih).

\begin{equation}
\begin{aligned}
\tau_{u_1} &\sim \mathit{Normal}_+(0,1)\\
\tau_{u_2} &\sim \mathit{Normal}_+(0,1)\\
\begin{bmatrix}
1 & \rho_u \\
\rho_u & 1
\end{bmatrix} \sim \mathit{LKJcorr}(2)
\end{aligned}
\end{equation}


We are now ready to fit the model. To speed up computation, we subset 50 subjects of the original data set; both the subsetted data and the original data set can be found in the package `bcogsci`. If we were analyzing these data for publication in a journal article or the like, we would obviously not subset the data.

We restrict ourselves to the correct trials only, and add a `c_cond` predictor, sum-coded as described earlier.

```{r stroopdata, message = FALSE}
data("df_stroop")
(df_stroop <- df_stroop %>%
  mutate(c_cond = if_else(condition == "Incongruent", 1, -1)))
```
Fit the model.

```{r stroopm1, message = FALSE, results = "hide", eval = !file.exists("dataR/fit_stroopm1.RDS")}
fit_stroop <- brm(RT ~ c_cond + (c_cond | subj),
                  family = lognormal(),
                  prior =
                    c(prior(normal(6, 1.5), class = Intercept),
                      prior(normal(0, .01), class = b),
                      prior(normal(0, 1), class = sigma),
                      prior(normal(0, 1), class = sd),
                      prior(lkj(2), class = cor)),
                  data = df_stroop)
```

```{r ,echo = FALSE}
if (!file.exists("dataR/fit_stroopm1.RDS")) {
  saveRDS(fit_stroop, "dataR/fit_stroopm1.RDS")
} else {
  fit_stroop <- readRDS("dataR/fit_stroopm1.RDS")
}
```

We will focus on $\beta$ (but you can verify that there is nothing surprising in the other parameters in the model `fit_stroop`).

```{r}
posterior_summary(fit_stroop, variable = "b_c_cond")
```

As shown in Figure \@ref(fig:priorposteriordiscrepancy),
if we overlay the density plots for the prior and posterior distributions of $\beta$, it becomes evident that the prior might have been too restrictive: the posterior is relatively far from the prior, and the prior strongly down-weights the values that the posterior is centered around. Such a strong discrepancy between the prior and posterior can be investigated with a sensitivity analysis.

(ref:priorposteriordiscrepancy) The discrepancy between the prior and the posterior distributions for the slope parameter in the model `fit_stroop`.

```{r priorposteriordiscrepancy,fig.cap="(ref:priorposteriordiscrepancy)", fig.height = 2.2}
sample_b_post <- as_draws_df(fit_stroop)$b_c_cond
# We generate samples from the prior as well:
N <- length(sample_b_post)
sample_b_prior <- rnorm(N, 0, .01)
samples <-
  tibble(sample = c(sample_b_post, sample_b_prior),
         distribution = c(rep("posterior", N), rep("prior", N)))
ggplot(samples, aes(x = sample, fill = distribution)) +
  geom_density(alpha = .5)
```

#### \index{Sensitivity analysis} Sensitivity analysis

Here, the discrepancy evident in Figure \@ref(fig:priorposteriordiscrepancy) is investigated with a sensitivity analysis. We will examine what happens for the following priors for $\beta$. In the models we fit below, all the other parameters have the same priors as in the model `fit_stroop`;  we vary only the priors for $\beta$.
The different priors are:

- $\beta \sim \mathit{Normal}(0,0.001)$
- $\beta \sim \mathit{Normal}(0,0.01)$
- $\beta \sim \mathit{Normal}(0,0.05)$
- $\beta \sim \mathit{Normal}(0,0.1)$
- $\beta \sim \mathit{Normal}(0,1)$
- $\beta \sim \mathit{Normal}(0,2)$

```{r stroopfits, message = FALSE, results = "hide", eval = !file.exists("dataR/df_beta_stroop.RDS"), echo = FALSE}
sds <- c(0.001, 0.01, 0.05, 0.1, 1, 2)
df_beta_stroop <- map_dfr(sds, function(sd) {
  priorb <- paste0("normal(0, ", sd, ")")
  fit <-
    brm(RT ~ c_cond + (c_cond | subj),
        family = lognormal(),
        prior =
          c(prior(normal(6, 1.5), class = Intercept),
            prior_string(priorb, class = "b"),
            prior(normal(0, 1), class = sigma),
            prior(normal(0, 1), class = sd),
            prior(lkj(2), class = cor)),
        iter = 4000,
        data = df_stroop)

  posterior_summary(fit, variable = "b_c_cond") %>%
    as_tibble() %>%
    mutate(prior = priorb)
})
```

```{r, echo = FALSE}
if (!file.exists("dataR//df_beta_stroop.RDS")) {
  saveRDS(df_beta_stroop, "dataR/df_beta_stroop.RDS")
} else {
  df_beta_stroop <- readRDS("dataR/df_beta_stroop.RDS")
}
```

We can summarize the estimates of $\beta$ given different priors as shown in Table \@ref(tab:priorsslopesmeansCIs).

```{r priorsslopesmeansCIs,echo = FALSE, results = "asis"}

if(knitr::is_html_output()) text  = "$(\\texttt{fit_stroop})$" else text  = "$(\\texttt{fit\\_stroop})$"

df_beta_stroop %>%
  select(-Est.Error) %>%
  mutate(prior =paste0("$", gsub("normal","\\\\mathit{Normal}", prior),"$")) %>%
  as.data.frame() %>%
  mutate(across(where(is.numeric), ~ paste0("$",formatC(round(.x, 3),format = "f", digits=3), "$"))) %>%
  select(prior, everything()) %>%
  kableExtra::kable(caption = paste0("The summary (mean and 95\\% credible interval) for the posterior distribution of the slope in the model ", text,", given different priors on the slope parameter."),
                    align =c("l","r","r","r"),
                    vline= "",
                    booktabs = TRUE,
                    escape = FALSE)
```



It might be easier to see how much the posterior difference between conditions changes depending on the prior. In order to answer this question, we need to remember that the median difference between conditions ($\mathit{MedianRT}_{\hbox{diff}}$) can be calculated as the difference between the exponents of each condition's medians:

\begin{equation}
\begin{aligned}
\mathit{MedianRT}_{\hbox{diff}} &= \mathit{MedianRT}_{\hbox{incongruent}} - \mathit{MedianRT}_{\hbox{congruent}}\\
\mathit{MedianRT}_{\hbox{diff}} &= \exp(\alpha + \beta) - \exp(\alpha - \beta)
\end{aligned}
(\#eq:medianrt)
\end{equation}

```{r stroopfitsdiff, message = FALSE, results = "hide", eval = !file.exists("_bookdown_files/df_diffs_stroop.RDS"), echo = FALSE}
sds <- c(0.001, 0.01, 0.05, 0.1, 1, 2)
df_diffs_stroop <- map_dfr(sds, function(sd)
{
  priorb <- paste0("normal(0, ", sd, ")")
  fit <- brm(RT ~ c_cond + (c_cond | subj),
             family = lognormal(),
             prior =
               c(prior(normal(6, 1.5), class = Intercept),
                 prior_string(priorb, class = "b"),
                 prior(normal(0, 1), class = sigma),
                 prior(normal(0, 1), class = sd),
                 prior(lkj(2), class = cor)),
             iter = 4000,
             data = df_stroop)
  sample_a_post <- as_draws_df(fit)$b_Intercept
  sample_b_post <- as_draws_df(fit)$b_c_cond
  RT_diff <- exp(sample_a_post + sample_b_post) -
    exp(sample_a_post - sample_b_post)
  tibble(prior = priorb,
         `mean diff (ms)` = mean(RT_diff),
         Q2.5 = quantile(RT_diff, .025),
         Q97.5 = quantile(RT_diff, .975))
})
```
```{r, echo = FALSE}
if (!file.exists("_bookdown_files/df_diffs_stroop.RDS")) {
  saveRDS(df_diffs_stroop, "dataR/df_diffs_stroop.RDS")
} else {
  df_diffs_stroop <- readRDS("dataR/df_diffs_stroop.RDS")
}
```

Equation \@ref(eq:medianrt) gives us the posterior distributions of the median difference between conditions for the different models. We calculate the median difference rather than the mean difference because the mean depends on the parameter $\sigma$, but the median doesn't: The mean of a log-normal distribution is $\exp(\mu +\sigma ^{2}/2)$, and the median is simply $\exp(\mu)$; see also \@ref(sec-lognormal). Another reason for reporting the median difference is that it would be a better way to summarize a skewed reading time or response time distribution (the mean is very sensitive to extreme values).

Table \@ref(tab:meanrtdiffsummary) summarizes the posterior distributions under different priors using the means of the difference in medians, along with 95% credible intervals. It's important to realize that the use of the mean to summarize the posterior distribution is orthogonal to our use of the median to summarize the response times by condition: In the first case, we use the median to summarize a group of *observations*, and in the second case, we use the mean to summarize a group of *samples* from the posterior--we could have summarized the samples from the posterior with its median instead of the mean.


```{r, meanrtdiffsummary,echo = FALSE, results = "asis"}
df_diffs_stroop %>%
   mutate(prior =paste0("$", gsub("normal","\\\\mathit{Normal}", prior),"$")) %>%
 as.data.frame() %>%
  mutate(across(where(is.numeric), ~ paste0("$",formatC(round(.x, 2),format = "f", digits=2), "$"))) %>%
  kableExtra::kable(escape = FALSE, align = c("l","r","r","r"),
                    vline= "", booktabs = TRUE,
                    caption="A summary, under a range of priors, of the posterior distributions of the mean difference between the two conditions, back-transformed to the millisecond scale.")
```

Table \@ref(tab:meanrtdiffsummary) shows us that the posterior changes substantially when we use wider priors. It seems that the posterior is relatively unaffected when we use priors with a standard deviation larger than $0.05$. However, if we assume a priori that the effect of the manipulation must be small, we will end up obtaining a posterior that is consistent with that belief.  When we include less information about the possible effect sizes by using a less informative prior, we allow the data to influence the posterior more. A sensitivity analysis is always an important component of a good-quality Bayesian analysis.

Which analysis should one report after carrying out a sensitivity analysis? In the above example, the priors ranging from $\mathit{Normal}(0,0.05)$ to $\mathit{Normal}(0,2)$ show rather similar posterior distributions for the mean difference. The most common approach in Bayesian analysis is to report the results of such relatively uninformative priors (e.g., one could report the posterior associated with the $\mathit{Normal}(0,2)$ here), because this kind of prior allows for a broader range of possible effects and is relatively agnostic. However, if there is a good reason to use a prior from a previous analysis, then of course it makes sense to report the analysis with the informative prior alongside an analysis with an uninformative prior. Reporting only informative priors in a Bayesian analysis is generally not a good idea. The issue is transparency: the reader should know what the posterior looks like for both an informative and an uninformative prior.

Another situation where posterior distributions associated with multiple priors should be reported is when one is carrying out an \index{Adversarial sensitivity analysis} adversarial sensitivity analysis [@spiegelhalter2004bayesian]: one can take a group of agnostic, enthusiastic, and adversarial or skeptical priors that, respectively, reflect a non-committal a priori position, an informed position based on the researcher's prior beliefs, and an adversarial position based on a scientific opponent's beliefs. In such a situation, analyses using all three priors can be reported, so that the reader can determine how different prior beliefs influence the posterior. For an example of such an adversarial analysis, see @VasishthEngelmann2022. Finally, when carrying out hypothesis testing using Bayes factors, the choice of the prior on the parameter of interest becomes critically important; in that situation, it is very important to report a sensitivity analysis, showing the Bayes factor as well as a summary of the posterior distributions [@SchadEtAlBF]; we return to this point in chapter \@ref(ch-bf), which covers Bayes factors.





## Why fitting a Bayesian hierarchical model is worth the effort

Carrying out Bayesian data analysis clearly requires much more effort than fitting a frequentist model: we have to define priors, verify that our model works, and decide how to interpret the results. By comparison, fitting a linear mixed model using \index{\texttt{lme4}} `lme4` consists of only a single line of code. But there is a hidden cost to the relatively high speed furnished by functions such as `lmer()`. First, the model fit using `lmer()` or the like  makes many assumptions, but they are hidden from the user. This is not a problem for the knowledgeable modeler, but very dangerous for the naive user. A second conceptual problem is that the way frequentist models are typically used is to answer a binary question: is the effect "significant" or not? If a result is significant, the paper is considered worth publishing; if not, it is not. Although frequentist models can quickly answer the question that the null hypothesis test poses, the frequentist test answers the wrong question. For discussion, see @VasishthNicenboimStatMeth.

Nevertheless, it is natural to ask why one should bother to go through all the trouble of fitting a Bayesian model. An important reason is the flexibility in model specification. The approach we have presented here can be used to extend essentially any parameter of any model. This includes popular uses, such as logistic and Poisson regressions, and also useful models that are relatively rarely used in cognitive science, such as multi-logistic regression [e.g., accuracy in some task with more than two answers), ordered logistic (e.g., ratings, @burkner2019ordinal], models with a shifted log-normal distribution [see online exercise \@ref(exr:shiftedlogn) and chapter \@ref(ch-lognormalrace) which deals with a log-normal race mode, and see @NicenboimEtAl2016Frontiersb; @Rouder2005], and distributional regression models (as shown in section \@ref(sec-distrmodel)). By contrast, a frequentist model, although easy to fit quickly, forces the user to use an inflexible canned model, which may not necessarily make sense for their data.

 This flexibility allows us also to go beyond the statistical models discussed before, and to develop complex hierarchical computational process models that are tailored to specific phenomena. An example are computational cognitive models, these can be extended hierarchically in a straightforward way, see @Lee2011 and @LeeWagenmakers2014. This is because, as we have seen with distributional regression models in section \@ref(sec-distrmodel), any parameter can have a group-level effect structure. Some examples of hierarchical computational cognitive models in psycholinguistics are @LogacevVasishth2015, @nicenboimModelsRetrievalSentence2018, @VasishthEtAl2017Modelling,  @VasishthEtAl2017Feature, @lisson_2020, @logacev-dokudan-2021-multinomial, @PaapeEtAlMPT2020, @yadavencret2021, and @yadaviccm2021. The hierarchical Bayesian modeling approach can even be extended to process models that cannot be expressed as a likelihood function, although in such cases one may have to write one's own sampler; for an example from psycholinguistics, see @yadavindiff2021.^[Most of the papers mentioned above provide example code using Stan or `brms`.] We discuss and implement in Stan some relatively simple computational cognitive models in chapters \@ref(ch-cogmod)-\@ref(ch-lognormalrace).

## Summary

This chapter presents two very commonly used classes of hierarchical model: those with normal and log-normal likelihoods. We saw several common variants of such models: varying intercepts, varying intercepts and varying slopes with or without a correlation parameter, and crossed random effects for subjects and items. We also experienced the flexibility of the Stan modeling framework through the example of a model that assumes a different residual standard deviation for each subject.

## Further reading

Chapter 5 of @Gelman14 provides a rather technical but complete treatment of exchangeability in Bayesian hierarchical models. @bernardosmith is a brief but useful article explaining exchangeability, and @lunn2012bugs also has a helpful discussion that we have drawn on in this chapter. @GelmanHill2007 is a comprehensive treatment of hierarchical modeling, although it uses WinBUGS.  @yarkoni_2020 discusses the importance of modeling variability in variables that researchers clearly intend to generalize over (e.g., stimuli, tasks, or research sites), and how under-specification of group-level effects imposes strong constraints on the generalizability of results. @SorensenVasishthTutorial provides an introduction, using Stan, to the Laird-Ware style matrix formulation [@laird1982random] of hierarchical models; this formulation has the advantage of flexibility and efficiency when specifying models in Stan syntax.

